---
output: github_document
---

<!-- README.md is generated from README.Rmd. Please edit that file -->

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.path = "man/figures/README-",
  out.width = "100%"
)
```

# smsR

<!-- badges: start -->
<!-- badges: end -->

smsR is a seasonal stock assessment model package using TMB to estimate parameters. The model is the current assessment model for sandeel (4 stocks) and sprat in the North Sea. 

## Installation

You can install the development version of smsR from [GitHub](https://github.com/nissandjac/smsR) with:

``` r
# install.packages("devtools")
devtools::install_github("nissandjac/smsR")
```

## Using the model

The model can be run with the provided data set `{r} sandeel_1r` and the Sandeel 1r standard configuration as 

```{r example}
library(smsR)


# Set up sandeel for area 1r 

df.tmb <- get_TMB_parameters(
  mtrx = sandeel_1r$lhs, # List that contains M, mat, west, weca
  Surveyobs = sandeel_1r$survey, # Survey observations 
  Catchobs = sandeel_1r$Catch, # Catch observations  
  years = 1983:2021, # Years to run
  nseason = 2, # Number of seasons
  useEffort = TRUE, # Use effort to calculate F
  ages = 0:4, # Ages of the species
  recseason = 2, # Season where recruitment occurs
  CminageSeason = c(1,1), # Minimum catch age per season
  Fmaxage = 3, # Fully selected fishing mortality age
  Qminage = c(0,1),# Minimum age in surveys 
  Qmaxage = c(1,3), # Max age in surveys 
  Fbarage = c(1,2), # Age use to calculate Fbar 
  effort = sandeel_1r$effort, # Effort input 
  blocks = c(1983,1999), # Blocks with unique selectivity 
  nocatch = sandeel_1r$nocatch, # Seasons where F is not calculated 
  surveyStart = c(0.75,0), #
  surveyEnd =  c(1,0), # 
  surveySeason = c(2,1), # 
  surveyCV =  list(c(0,1),c(1,2)), 
  catchCV = list(c(1,3), c(1,3)), # Catch CV groupings 
  estCV = c(0,2,0), # Estimate CVs for 1) survey, 2) catch, 3) Stock recruitment relationship
  beta = 105809, # Hockey stick break point
  nllfactor = c(1,1,0.05) # Factor for relative strength of log-likelihood

)


parms <- getParms(df.tmb)

sas <- runAssessment(df.tmb, parms)


```



<!-- You'll still need to render `README.Rmd` regularly, to keep `README.md` up-to-date. `devtools::build_readme()` is handy for this. -->

The resulting fit to the data can then be visualized in a summary output showing SSB, Catch, recruitment and Fbar as 

```{r, echo = TRUE}

plot(sas, printFig = FALSE)

```


Or to evaluate the retrospective patterns as 

```{r, echo = TRUE, warning=FALSE, message=FALSE}

mr <- mohns_rho(df.tmb, peels = 5, parms = parms, plotfigure = FALSE)

mr$p1()

```


